Spatial Reasoning and Statistical Graphics

### 2014-11-14 Paper/Draft1/TestingVisualAptitude.Rnw
few edits, nothing major

\comment{Next few steps for the analysis: \\
(1) I think we should add a linear model for lineup scores that includes the demographics of participants as well; \\
\textbf{Done; added to Results section}\\
(2) could you do the scoring of lineups by type of plot used and investigate the relationship between those types of lineups and the other visual tests ... not sure that that will pay off, but it's interesting.\\
\textbf{Done, along with an examination of each experiment separately}\\
(3) in terms of why we would use lineups in a test comparison ... lineups give us a way to say 'this is correct' which regular graphics don't. So we should put a spin sentence on at the intro and the back that we are suggesting to use lineups as a tool to not only evaluate individuals (which we can see is possible) but also as a tool to evaluate charts ... which we've done before ... and tie it into the regular testing literature ...\\
\textbf{A few sentences added, still need to tie into the existing literature}}

\todo[inline]{Intro about statistical graphics, lineups, ...}
\subsection{Spatial Reasoning and Statistical Graphics}
Statistical graphics provide quick summaries of data, models, and results, but these displays may not be equally useful to all viewers. Mathematical ability is important, even for the simplest graphs:  mathematical ability, not spatial ability, was shown~\cite{shah1995conceptual} to be associated with accuracy on a simple two-dimensional line graph. Spatial ability becomes more important, however, when more complicated graphical displays are used in comparison tasks: the lower performance of individuals with low spatial ability on tests utilizing diagrams and graphs is attributed ~\cite{mayer1994whom} to the fact that more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. It is theorized that graphics are a form of ``external cognition"~\cite{scaife1996external} that guide, constrain, and facilitate cognitive behavior\cite{zhang1997nature}. 
\newtext{Graphics utilize higher-bandwidth visual pathways to encode information\cite{baddeley1974working}, allowing viewers to quickly and intuitively encode multiple dimensions of numerical relationships; however, this higher-bandwidth approach also depends on the viewer's visual and spatial reasoning ability. Well-designed graphics present the most important features of the data while minimizing less-important features, guiding the viewer towards conclusions that are meaningful in context and supported by the data while maximizing the information encoded in working memory. Under this framework, well-designed graphics reduce memory load and make more cognitive resources available for other tasks (such as drawing conclusions from the data), at the cost of depending on certain visuospatial reasoning abilities. }

@@

\subsection{The Lineup Protocol}
\todo[inline]{Standard lineup protocol introduction here.} 
\newtext{Figure \ref{fig:lineup} shows a sample lineup of boxplots; participants would be expected to determine that sub-plot 4 is the most different because the two groups have notably different medians.}

@@

\newtext{One of the most useful features of the lineup protocol is that it allows researchers to conclusively determine which graphics show certain features most conclusively (providing an experimental protocol for comparing graphics based on the accuracy of user conclusions). In addition, lineups provide researchers with a statistically valid methodology for determining whether a specific graph shows a real, statistically significant effect by comparing a target plot with plots formed using permutations of the same data, providing a randomization test protocol for graphics. As a result, lineups are a useful and innovative tool for evaluating charts; on an individual level, they can also be used to evaluate a specific participant's perceptual reasoning ability in the context of statistical graphics.}

>>>>

\comment{Next few steps for the analysis: \\
\\
(3) in terms of why we would use lineups in a test comparison ... lineups give us a way to say 'this is correct' which regular graphics don't. So we should put a spin sentence on at the intro and the back that we are suggesting to use lineups as a tool to not only evaluate individuals (which we can see is possible) but also as a tool to evaluate charts ... which we've done before ... and tie it into the regular testing literature ...\\
\textbf{A few sentences added, still need to tie into the existing literature}
(4) I'll take care of the intro to lineups}

\todo[inline]{Intro about statistical graphics, lineups, ...}
\subsection{Spatial Reasoning and Statistical Graphics}
Statistical graphics provide quick summaries of data, models, and results, but these displays may not be equally useful to all viewers. Mathematical ability is important, even for the simplest graphs:  mathematical ability, not spatial ability, was shown~\cite{shah1995conceptual} to be associated with accuracy on a simple two-dimensional line graph. Spatial ability becomes more important, however, when more complicated graphical displays are used in comparison tasks: the lower performance of individuals with low spatial ability on tests utilizing diagrams and graphs is attributed ~\cite{mayer1994whom} to the fact that more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. It is theorized that graphics are a form of ``external cognition"~\cite{scaife1996external} that guide, constrain, and facilitate cognitive behavior\cite{zhang1997nature}. 
Graphics utilize higher-bandwidth visual pathways to encode information\cite{baddeley1974working}, allowing viewers to quickly and intuitively encode multiple dimensions of numerical relationships; however, this higher-bandwidth approach also depends on the viewer's visual and spatial reasoning ability. Well-designed graphics emphasize and present important features of the data while minimizing  features of lesser importance, guiding the viewer towards conclusions that are meaningful in context and supported by the data while maximizing the information encoded in working memory. Under this framework, well-designed graphics reduce memory load and make more cognitive resources available for other tasks (such as drawing conclusions from the data), at the cost of depending on certain visuospatial reasoning abilities.

@@

\subsection{The Lineup Protocol}
\todo[inline]{Standard lineup protocol introduction here.}

\newtext{Figure \ref{fig:lineup} shows a sample lineup of boxplots; participants would be expected to determine that sub-plot 4 is the most different because the two groups have notably different medians.}

@@

\newtext{One of the most useful features of the lineup protocol is that it allows researchers to conclusively determine which graphics show certain features more conclusively (providing an experimental protocol for comparing graphics based on the accuracy of user conclusions). In addition, lineups provide researchers with a rigorous framework for determining whether a specific graph shows a real, statistically significant effect by comparing a target plot with plots formed using permutations of the same data, providing a randomization test protocol for graphics. As a result, lineups are a useful and innovative tool for evaluating charts; on an individual level, they can also be used to evaluate a specific participant's perceptual reasoning ability in the context of statistical graphics.}
